---
title: "analysis"
author: "Will Oestreich"
date: '2025-02-18'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, # show warnings
  message = FALSE, # show messages
  error = TRUE, # do not interrupt generation in case of errors,
  echo = TRUE  # show R code
)
```

```{r}
library(tidyverse) 
library(dplyr)
library(sf) 
library(terra)
library(raster)
library(arrow)
library(stringr)
library(MetBrewer)
library(rnaturalearth)
library(rnaturalearthdata)
library(scales)
library(cowplot)
library(fst)
```

## load in parquet
```{r}
obis_snapshot <- "data/obis_20241202.parquet"
```

## open dataset
```{r}
db <- open_dataset(obis_snapshot)
```

## quality filters
```{r}
obis_filter <- db |>
  filter(!str_detect(flags, "DATE_IN_FUTURE")) |> 
  filter(!str_detect(flags, "DATE_BEFORE_MIN")) |> 
  filter(!str_detect(flags, "NO_COORD")) |> 
  filter(!str_detect(flags, "ZERO_COORD")) |> 
  filter(!str_detect(flags, "LON_OUT_OF_RANGE")) |> 
  filter(!str_detect(flags, "LAT_OUT_OF_RANGE")) |> 
  filter(!str_detect(flags, "NO_DEPTH")) |> 
  filter(!str_detect(flags, "ON_LAND")) |> 
  mutate(bath_obs_prop_min = minimumDepthInMeters/bathymetry) |>
  mutate(bath_obs_prop_max = maximumDepthInMeters/bathymetry) |>
  mutate(bathymetry = round(as.numeric(bathymetry), 0)) |>
  mutate(minimumDepthInMeters = round(as.numeric(minimumDepthInMeters), 0)) |>
  mutate(depth_diff = bathymetry - minimumDepthInMeters) |>
  mutate(depth_quality_category = ifelse(bath_obs_prop_min <= 1.05 & bath_obs_prop_min >= 0, "depth_good", "depth_bad")) |>
  mutate(latitude = round(decimalLatitude,3), 
         longitude = round(decimalLongitude,3)) |>
  group_by(bathymetry, minimumDepthInMeters, bath_obs_prop_min, bath_obs_prop_max, depth_quality_category, latitude, longitude) |>
  count() |> 
  collect()
```

## explore outcome of custom depth quality flag
```{r}
depth_category_sums <- obis_filter |>
  group_by(depth_quality_category) |>
  summarize(total_n = sum(n))
head(depth_category_sums)
```

## filter out records with bad depth vs. bathymetry comparison flag
```{r}
obis_clean <- obis_filter |> 
  filter(depth_quality_category == "depth_good")
```

## the custom flag lets in records with depth up to 5% greater than bottom depth. for these records, assign depth to bottom depth and make the proportional depth = 1
```{r}
obis_clean <- obis_clean |>
  mutate(bath_obs_prop_min = ifelse(bath_obs_prop_min > 1, 1, bath_obs_prop_min)) |>
  mutate(minimumDepthInMeters = ifelse(bath_obs_prop_min > 1, bathymetry, minimumDepthInMeters))
```

## depth zones
```{r}
obis_clean <- obis_clean %>%
  mutate(zone = case_when(
    bathymetry < 200 ~ "coastal or shallow",
    bathymetry >= 200 & minimumDepthInMeters < 200 ~ "coastal or shallow",
    bathymetry >= 200 & minimumDepthInMeters > 200  ~ "deep",
  ))
```

## summarize by depth zone
```{r}
obis_zone_summary <- obis_clean %>%
  filter(!is.na(zone)) %>%  
  group_by(zone) %>%
  summarise(total_n = sum(n, na.rm = TRUE))

obis_zone_summary$pct <- obis_zone_summary$total_n/sum(obis_zone_summary$total_n, na.rm = TRUE)
head(obis_zone_summary)
```

## save cleaned OBIS dataset and depth zone summary
```{r}
write.csv(obis_zone_summary,"outputs/obis_zone_summary.csv")
write_fst(obis_clean,"outputs/obis_clean.fst",compress = 90)
```

## FIGURE PANEL A: maps - centered on international date line
```{r}
df_spatial <- obis_clean |> 
  dplyr::filter(!is.na(latitude), 
         !is.na(longitude)) |> 
  st_as_sf(coords = c("longitude", "latitude"), 
           crs = "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m")
```

```{r}
zone_list <- split(df_spatial, df_spatial$zone)

raster_list <- list()

crs <- "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m" 
for (zone_name in names(zone_list)) {
  zone_data <- zone_list[[zone_name]]
  
  ras_temp <- raster(xmn=-180, xmx=180, ymn=-90, ymx=90, resolution=c(0.1, 0.1), vals=NA)
  raster_zone <- rasterize(zone_data, ras_temp, field = "n", fun = 'sum')
  
  terra_zone <- terra::rast(raster_zone)
  terra_zone_proj <- terra::project(terra_zone, crs, mask=TRUE)
  
  raster_list[[zone_name]] <- terra_zone_proj
}

raster_stack <- terra::rast(raster_list)
```

```{r} 
rob_proj <- "+proj=robin +lon_0=180"  # Change to Pacific-centric projection
raster_stack_robin <- project(raster_stack, rob_proj)

world <- ne_countries(scale = "medium", returnclass = "sf")
## attempt to fix meridian issues
polygon <- st_polygon(x = list(rbind(c(-0.0001, 90),
                                     c(0, 90),
                                     c(0, -90),
                                     c(-0.0001, -90),
                                     c(-0.0001, 90)))) %>%
  st_sfc() %>%
  st_set_crs(4326)

world2 <- world %>% st_difference(polygon)
world_robinson <- st_transform(world2, 
                               crs = '+proj=robin +lon_0=180 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs')

# Custom colormap
colors <- c("#4491B2", "#306F9F", "#1C4772", "#102F4F")

# Set the colorbar limits
color_limits <- c(0, 6)

# Loop through each raster layer and create/save a separate plot
for (i in 1:nlyr(raster_stack_robin)) {
  # Extract the current layer as a data frame
  raster_df <- as.data.frame(raster_stack_robin[[i]], xy = TRUE)
  colnames(raster_df) <- c("x", "y", "value")  # Standardize column names
  
  # Get the layer name
  layer_name <- names(raster_stack_robin)[i]
  
  # Create the plot
   plot <- ggplot() +
    geom_raster(data = raster_df, aes(x = x, y = y, fill = log10(value))) +
    geom_sf(data = world_robinson, fill = "grey", color = NA) +
    scale_fill_gradientn(
      colors = colors, 
      name = expression(log[10]("observations")),  # Add a label for the colorbar
      limits = color_limits,
      guide = "none"
    ) +
    coord_sf(crs = rob_proj) +
    theme_minimal() +
    theme(
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      panel.grid = element_blank(),
      legend.position = "top",
      legend.text = element_text(size = 24),  # Adjust font size for legend labels
      legend.title = element_text(size = 24)  # Adjust font size for the legend title
    )
  
  # Display the plot 
  print(plot)
  
  # Save the plot to a file
  ggsave(filename = paste0("figures/", layer_name, "_map.png"), plot = plot, width = 10, height = 6, dpi = 300)
}
```

## FIGURE PANEL B: depth histogram
```{r} 
p <- obis_clean |>
  ggplot(aes(x = minimumDepthInMeters, weight = n)) +
  geom_histogram(binwidth = 100, fill = "#306F9F") +
  theme_bw() +
  coord_flip() +
  scale_x_reverse() +
  labs(x = "Depth (m)", y = "Biodiversity observations") +
  geom_vline(xintercept = 200, linetype = "dashed", color = "black")

ggsave("figures/depth.png",p,width=2.3,height=5.2,units="in")
```

## FIGURE PANEL C: pie charts of biodiversity observations and habitable volume
```{r} 
zone_colors <- c("deep" = "#1C4772",
                 "coastal or shallow" = "#4491B2",
                 "land" = "#156E33")

# Pie chart for biodiversity observations
pa <- ggplot(pie, aes(x = "", y = pct, fill = zone)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = zone_colors) +
  theme_void() +
  theme(legend.position = "none")

# Pie chart for habitable space by volume
pb <- ggplot(pie, aes(x = "", y = space, fill = zone)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = zone_colors) +
  theme_void() +
  theme(legend.position = "none")

ggsave("figures/pie_a.png",pa)
ggsave("figures/pie_b.png",pb)
```

## FIGURE PANEL D: observations by node, classified by income level
```{r}
nodes_classification <- read_csv("data/nodes_obs_classified.csv") |>
  dplyr::select(node_name, n, class)

node_obs <- nodes_classification |>
  group_by(class) |>
  summarise(n = sum(n)) |>
  arrange(-n) |>
  mutate(all = sum(n)) |>
  mutate(perc = n/all*100) |>
  arrange(-perc)

node_obs$class <- factor(
  node_obs$class,
  levels = c("Not classified", "Low income","Lower middle income","Upper middle income","High income"))

p <- ggplot(node_obs, aes(x = class, y = perc)) + geom_col(fill = "#306F9F") + coord_flip() +
  ylab("Biodiversity observations\nmanaged (%)") +
  xlab("") +
  ylim(c(0,100)) +
  theme_bw() +
  theme(axis.text.y = element_text(angle = 45, hjust = 1))

ggsave("figures/nodes.png",p,width=3,height=3.3,units="in")
```

